RProtoBuf: Efficient Cross-Language Data Serialization in R

### 2014-01-03 papers/rjournal/eddelbuettel-francois-stokely.Rnw
Improve section 2 on protocol buffers.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@698 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Given these requirements, how do we safely share intermediate results
between different applications, possibly written in different
languages, and possibly running on different computers?  Programming
languages such as R, Java, Julia, and Python include built-in
serialization support, but these formats are tied to the specific
programming language in use and thus lock the user into a single
environment.  CSV files can be read and written by many applications
and so are often used for exporting tabular data.  However, CSV files
have a number of disadvantages, such as a limitation of exporting only
tabular datasets, lack of type-safety, inefficient text representation
and parsing, and ambiguities in the format involving special
characters.  JSON is another widely-supported format used mostly on
the web that removes many of these disadvantages, but it too suffers
from being too slow to parse and also does not provide strong typing
between integers and floating point.  Because the schema information
is not kept separately, multiple JSON messages of the same type
needlessly duplicate the field names with each message.
%
%
%
A number of binary formats based on JSON have been proposed that
reduce the parsing cost and improve the efficiency.  MessagePack
\citep{msgpackR} and BSON \citep{rmongodb} both have R interfaces, but
these formats lack a separate schema for the serialized data and thus
still duplicate field names with each message sent over the network or
stored in a file.  Such formats also lack support for versioning when
data storage needs evolve over time.

Once the data serialization needs of an application become complex
enough, developers typically benefit from the use of an
\emph{interface description language}, or \emph{IDL}.  IDLs like
Google's Protocol Buffers, Apache Thrift, and Apache Avro provide a compact
well-documented schema for cross-langauge data structures and
efficient binary interchange formats.  The schema can be used to
generate model classes for statically typed programming languages such
as C++ and Java, or can be used with reflection for dynamically typed
programming languages.  Since the schema is provided separately from
the encoded data, the data can be efficiently encoded to minimize
storage costs of the stored data when compared with simple
``schema-less'' binary interchange formats.

@@

This article describes the basics of Google's Protocol Buffers through
an easy to use R package, \CRANpkg{RProtoBuf}.  After describing the
basics of protocol buffers and \CRANpkg{RProtoBuf}, we illustrate
several common use cases for protocol buffers in data analysis.

\section{Protocol Buffers}
% This content is good.  Maybe use and cite?
% http://martin.kleppmann.com/2012/12/05/schema-evolution-in-avro-protocol-buffers-thrift.html

Protocol Buffers are a widely used modern language-neutral, platform-neutral, extensible mechanism for sharing structured data.

one of the more popular examples of the modern

XXX Related work on IDLs (greatly expanded )

XXX Design tradeoffs: reflection vs proto compiler

% TODO(ms) Also talk about versioning and why its useful.

%BSON, msgpack, Thrift, and Protocol Buffers take this latter approach,
%with the

% There are references comparing these we should use here.

TODO Also mention Thrift and msgpack and the references comparing some
of these tradeoffs.

@@

%% TODO(de,ms)  What follows is oooooold and was lifted from the webpage
%%              Rewrite?
Protocol buffers are a language-neutral, platform-neutral, extensible
way of serializing structured data for use in communications
protocols, data storage, and more.

Protocol Buffers offer key features such as an efficient data interchange
format that is both language- and operating system-agnostic yet uses a
lightweight and highly performant encoding, object serialization and
de-serialization as well data and configuration management. Protocol
buffers are also forward compatible: updates to the \texttt{proto}
files do not break programs built against the previous specification.

While benchmarks are not available, Google states on the project page that in
comparison to XML, protocol buffers are at the same time \textsl{simpler},
between three to ten times \textsl{smaller}, between twenty and one hundred
times \textsl{faster}, as well as less ambiguous and easier to program.

The protocol buffers code is released under an open-source (BSD) license. The
protocol buffer project (\url{http://code.google.com/p/protobuf/})
contains a C++ library and a set of runtime libraries and compilers for
C++, Java and Python.

With these languages, the workflow follows standard practice of so-called
Interface Description Languages (IDL)
(c.f. \href{http://en.wikipedia.org/wiki/Interface_description_language}{Wikipedia
  on IDL}).  This consists of compiling a protocol buffer description file
(ending in \texttt{.proto}) into language specific classes that can be used
to create, read, write and manipulate protocol buffer messages. In other
words, given the 'proto' description file, code is automatically generated
for the chosen target language(s). The project page contains a tutorial for
each of these officially supported languages:
\url{http://code.google.com/apis/protocolbuffers/docs/tutorials.html}

Besides the officially supported C++, Java and Python implementations, several projects have been
created to support protocol buffers for many languages. The list of known
languages to support protocol buffers is compiled as part of the
project page: \url{http://code.google.com/p/protobuf/wiki/ThirdPartyAddOns}

@@

The two fundamental building blocks of Protocol Buffers are Messages
and Descriptors.  Messages provide a common abstract encapsulation of
structured data fields of the type specified in a Message Descriptor.
Message Descriptors are defined in \texttt{.proto} files and define a
schema for a particular named class of messages.  This separation
between schema and the message objects is in contrast to
more verbose formats like JSON, and when combined with the efficient
binary representation of any Message object explains a large part of
the performance and storage-space advantage offered by Protocol
Buffers. TODO(ms): we already said some of this above.  clean up.

@@

@@

TODO(ms): random citations to work in:

>>>>

Given these requirements, how do we safely share intermediate results
between different applications, possibly written in different
languages, and possibly running on different computers?  Programming
languages such as R, Julia, Java, and Python include built-in
serialization support, but these formats are tied to the specific
programming language in use and thus lock the user into a single
environment.  CSV files can be read and written by many applications
and so are often used for exporting tabular data.  However, CSV files
have a number of disadvantages, such as a limitation of exporting only
tabular datasets, lack of type-safety, inefficient text representation
and parsing, and ambiguities in the format involving special
characters.  JSON is another widely-supported format used mostly on
the web that removes many of these disadvantages, but it too suffers
from being too slow to parse and also does not provide strong typing
between integers and floating point.  Because the schema information
is not kept separately, multiple JSON messages of the same type
needlessly duplicate the field names with each message.
%
%
%
A number of binary formats based on JSON have been proposed that
reduce the parsing cost and improve the efficiency.  MessagePack
\citep{msgpackR} and BSON \citep{rmongodb} both have R interfaces, but
these formats lack a separate schema for the serialized data and thus
still duplicate field names with each message sent over the network or
stored in a file.  Such formats also lack support for versioning when
data storage needs evolve over time.

Once the data serialization needs of an application become complex
enough, developers typically benefit from the use of an
\emph{interface description language}, or \emph{IDL}.  IDLs like
Protocol Buffers \citep{protobuf}, Apache Thrift, and Apache Avro provide a compact
well-documented schema for cross-langauge data structures and
efficient binary interchange formats.  The schema can be used to
generate model classes for statically typed programming languages such
as C++ and Java, or can be used with reflection for dynamically typed
programming languages.  Since the schema is provided separately from
the encoded data, the data can be efficiently encoded to minimize
storage costs of the stored data when compared with simple
``schema-less'' binary interchange formats.

@@

% TODO(mstokely): If we go to JSS, include a larger paragraph here
% referencing each numbered section.  I don't like these generally,
% but its useful for this paper I think because we have a boring bit
% in the middle (full class/method details) and interesting
% applications at the end.
This article describes the basics of Google's Protocol Buffers through
an easy to use R package, \CRANpkg{RProtoBuf}.  After describing the
basics of protocol buffers and \CRANpkg{RProtoBuf}, we illustrate
several common use cases for protocol buffers in data analysis.

\section{Protocol Buffers}

@@

% This content is good.  Maybe use and cite?
% http://martin.kleppmann.com/2012/12/05/schema-evolution-in-avro-protocol-buffers-thrift.html

%% TODO(de,ms)  What follows is oooooold and was lifted from the webpage
%%              Rewrite?
Protocol Buffers are a modern language-neutral, platform-neutral,
extensible mechanism for sharing and storing structured data.  They
have been widely adopted in industry with applications as varied as Sony
Playstations, Twitter, Google Search, Hadoop, and Open Street Map.  While
traditional IDLs were previously characterized by bloat and
complexity, Protocol Buffers is based on a simple list and records
model that is flexible and easy to use.  Some of the key features
provided by Protocol Buffers for data analysis include:

\begin{itemize}
\item \emph{Portable}:  Allows users to send and receive data between
  applications or different computers.
\item \emph{Efficient}:  Data is serialized into a compact binary
  representation for transmission or storage.
\item \emph{Exentsible}:  New fields can be added to Protocol Buffer Schemas
  in a forward-compatible way that do not break older applications.
\item \emph{Stable}:  Protocol Buffers have been in wide use for over a
  decade.
\end{itemize}

Figure~\ref{fig:protobuf-distributed-usecase} illustrates an example
communication workflow with protocol buffers and an interactive R
session.  Common use cases include populating a request RPC protocol
buffer in R that is then serialized and sent over the network to a
remote server.  The server would then deserialize the message, act on
the request, and respond with a new protocol buffer over the network.

%Protocol buffers are a language-neutral, platform-neutral, extensible
%way of serializing structured data for use in communications
%protocols, data storage, and more.

%Protocol Buffers offer key features such as an efficient data interchange
%format that is both language- and operating system-agnostic yet uses a
%lightweight and highly performant encoding, object serialization and
%de-serialization as well data and configuration management. Protocol
%buffers are also forward compatible: updates to the \texttt{proto}
%files do not break programs built against the previous specification.

%While benchmarks are not available, Google states on the project page that in
%comparison to XML, protocol buffers are at the same time \textsl{simpler},
%between three to ten times \textsl{smaller}, between twenty and one hundred
%times \textsl{faster}, as well as less ambiguous and easier to program.

%The flexibility of the reflection-based API is particularly well
%suited for interactive data analysis.

% XXX Design tradeoffs: reflection vs proto compiler

For added speed and efficiency, the C++, Java, and Python bindings to
Protocol Buffers are used with a compiler that translates a protocol
buffer schema description file (ending in \texttt{.proto}) into
language-specific classes that can be used to create, read, write and
manipulate protocol buffer messages.  The R interface, in contrast,
uses a reflection-based API that is particularly well suited for
interactive data analysis.  All messages in R have a single class
structure, but different accessor methods are created at runtime based
on the name fields of the specified message type.

% In other words, given the 'proto'
%description file, code is automatically generated for the chosen
%target language(s). The project page contains a tutorial for each of
%these officially supported languages:
%\url{http://code.google.com/apis/protocolbuffers/docs/tutorials.html}

%The protocol buffers code is released under an open-source (BSD) license. The
%protocol buffer project (\url{http://code.google.com/p/protobuf/})
%contains a C++ library and a set of runtime libraries and compilers for
%C++, Java and Python.

%With these languages, the workflow follows standard practice of so-called
%Interface Description Languages (IDL)
%(c.f. \href{http://en.wikipedia.org/wiki/Interface_description_language}{Wikipedia
%  on IDL}).  This consists of compiling a protocol buffer description file
%(ending in \texttt{.proto}) into language specific classes that can be used

%Besides the officially supported C++, Java and Python implementations, several projects have been
%created to support protocol buffers for many languages. The list of known
%languages to support protocol buffers is compiled as part of the
%project page: \url{http://code.google.com/p/protobuf/wiki/ThirdPartyAddOns}

@@

The two fundamental building blocks of Protocol Buffers are Messages
and Descriptors.  Messages provide a common abstract encapsulation of
structured data fields of the type specified in a Message Descriptor.
Message Descriptors are defined in \texttt{.proto} files and define a
schema for a particular named class of messages.

@@

% Commented out because we said this earlier.
%This separation
%between schema and the message objects is in contrast to
%more verbose formats like JSON, and when combined with the efficient
%binary representation of any Message object explains a large part of
%the performance and storage-space advantage offered by Protocol
%Buffers. TODO(ms): we already said some of this above.  clean up.

@@

% RProtoBuf has been used.
